FHJ: A Formal Model for Hierarchical Dispatching and Overriding

### 2018-06-11 ECOOP18/sections/Introduction.tex
revision in progress
remove line number, comments, compress to 25pages.

In contrast to diamond inheritance, the second case of ambiguity
is \textit{unintentional method conflicts}~\cite{scharli03traits}. In
this case, conflicting 
methods do not actually refer to the same feature. 
In a nominal system, methods can be designed for different
functionality, but happen to have the same names (and signatures).
A simple example of this situation is two \lstinline{draw} methods that
are inherited from a deck of cards and a drawable widget, respectively. 
In such context, the two \lstinline{draw} methods have very different meanings, 
but they happen to share the same name.
%%This issue was proposed by the trait paper, so-called
When inheritance is used to compose these classes, a compilation 
error happens due to conflicts. However, unlike the diamond problem,
the conflicting methods have very different meanings and do not share a
common parent. We call such a case \textit{\wordfork{} inheritance}, in
analogy to diamond inheritance.

@@

\textit{Hierarchical overriding} is a novel language
mechanism that allows method overriding to be applied
only to one branch of the class hierarchy. Hierarchical overriding 
adds expressive power that is not available in languages such as 
C++ or C\#. Hierarchical overriding allows overriding to work 
for classes with multiple (conflicting) methods sharing the 
same names and signatures. An example is presented in 
Figure~\ref{fig:main}. In this example, there are 4 classes/interfaces.
Two classes \lstinline|Deck| and \lstinline|Drawable| model a deck of cards
 and a drawable widget, respectively. 
The class \lstinline|SafeDeck| adds functionality to check whether the
deck is empty so as to prevent drawing a card from an empty deck. 
The interesting class is \lstinline|DrawableSafeDeck|, which inherits 
from both \lstinline|SafeDeck| and \lstinline|Drawable|. Hierarchical
overriding is used in \lstinline|DrawableSafeDeck| to keep two separate \lstinline|draw|
methods for each parent, but override \emph{only} the \lstinline|draw|
method coming from \lstinline|Drawable|, in order to draw a widget
with a deck of cards. Note that hierarchical overriding is denoted in the UML
diagram with the notation
\lstinline|draw()|$\uparrow$\lstinline|Drawable|, expressing that the \lstinline|draw| method from \lstinline|Drawable|
is overridden. Although in this example only one of the
\lstinline|draw| methods is overridden (and the other is simply inherited), hierarchical overriding
supports multiple conflicting methods to be independently overridden as well.

>>>>

In contrast to diamond inheritance, the second case of ambiguity
is \textit{unintentional method conflicts}~\cite{scharli03traits}. In
this case, conflicting 
methods do not actually refer to the same feature. 
In a nominal system, methods can be designed for different
functionality but happen to have the same names (and signatures).
A simple example of this situation is two \lstinline{draw} methods that
are inherited from a deck of cards and a drawable widget, respectively. 
In such context, the two \lstinline{draw} methods have very different meanings, 
but they happen to share the same name.
%%This issue was proposed by the trait paper, so-called
When inheritance is used to compose these classes, a compilation 
error happens due to conflicts. However, unlike the diamond problem,
the conflicting methods have very different meanings and do not share a
common parent. We call such a case \textit{\wordfork{} inheritance}, in
analogy to diamond inheritance.

@@

\textit{Hierarchical overriding} is a novel language
mechanism that allows method overriding to be applied
only to one branch of the class hierarchy. Hierarchical overriding 
adds expressive power that is not available in languages such as 
C++ or C\#. In particular, it allows overriding to work 
for classes with multiple (conflicting) methods sharing the 
same names and signatures. An example is presented in 
Figure~\ref{fig:main}. In this example, there are 4 classes/interfaces.
Two classes \lstinline|Deck| and \lstinline|Drawable| model a deck of cards
 and a drawable widget, respectively. 
The class \lstinline|SafeDeck| adds functionality to check whether the
deck is empty so as to prevent drawing a card from an empty deck. 
The interesting class is \lstinline|DrawableSafeDeck|, which inherits 
from both \lstinline|SafeDeck| and \lstinline|Drawable|. Hierarchical
overriding is used in \lstinline|DrawableSafeDeck| to keep two separate \lstinline|draw|
methods for each parent, but override \emph{only} the \lstinline|draw|
method coming from \lstinline|Drawable|, in order to draw a widget
with a deck of cards. Note that hierarchical overriding is denoted in the UML
diagram with the notation
\lstinline|draw()|$\uparrow$\lstinline|Drawable|, expressing that the \lstinline|draw| method from \lstinline|Drawable|
is overridden. Although in this example only one of the
\lstinline|draw| methods is overridden (and the other is simply inherited), hierarchical overriding
supports multiple conflicting methods to be independently overridden as well.

### 2018-06-11 ECOOP18/sections/Overview.tex
revision in progress
remove line number, comments, compress to 25pages.

\begin{lstlisting}
interface DrawableSafeDeck extends Drawable, SafeDeck {}
// main program
new DrawableSafeDeck().shuffleAndDraw()
\end{lstlisting}

@@

% \begin{equation}\label{}
% \begin{split}
% A & = \frac{\pi r^2}{2} \\
%  & = \frac{1}{2} \pi r^2
% \end{split}
% \end{equation}

\begin{lstlisting}[style=reduction, escapechar=\%]
   %new DrawableSafeDeck()%.shuffleAndDraw()
-> %new DrawableSafeDeck().shuffle()%; new DrawableSafeDeck().draw()
-> ...
-> %new DrawableSafeDeck().draw()%
-> <<error: ambiguous call!!!>>
\end{lstlisting}

@@

\subsubsection{\MIM's solution} 
The essential problem is how to ensure that the correct method is
invoked. To solve this problem, \MIM{} uses a variant of method
dispatching that we call \textit{hierarchical dispatching}. In
hierarchical dispatching, both the static and dynamic type information 
are used to select the right method implementation.
%Inspired by C++, we have formalized this feature into a minimal core calculus,
%studied and proved the type soundness. It is concise and can readily be imported to other systems. 
%And in the formalization, we propose \textit{hierarchical dispatch} in our model
%as the default method lookup algorithm. 
During runtime, a method call
makes use of both the static type and the dynamic type of the
receiver, so it is a
combination of static and dynamic dispatching. Intuitively, the static type specifies one branch
to avoid ambiguity, and the dynamic type finds the most specific implementation on that branch.
To be specific, the following code is accepted by \MIM{}:
\begin{lstlisting}
interface Deck {
  void draw() {...}
  void shuffle() {...}
  void shuffleAndDraw() { this.shuffle(); this.draw(); }
}
interface Drawable {...}
interface SafeDeck extends Deck {...}
interface DrawableSafeDeck extends Drawable, SafeDeck {}
// main program
new DrawableSafeDeck().shuffleAndDraw() // SafeDeck.draw is called
\end{lstlisting}

\noindent The computation performed in \MIM{} is as follows:
\begin{lstlisting}[style=reduction, escapechar=\%]
   %new DrawableSafeDeck()%.shuffleAndDraw()
-> %((DrawableSafeDeck) new DrawableSafeDeck()).shuffleAndDraw()%
-> %((Deck) new DrawableSafeDeck()).shuffle()%; ((Deck) new DrawableSafeDeck()).draw()
-> ...
-> %((Deck) new DrawableSafeDeck()).draw()%
-> ... %// SafeDeck.draw%
\end{lstlisting}

@@

\subsubsection{\MIM{}'s solution} An additional feature of our model is \textit{hierarchical overriding}. It allows conflicting methods
to be overridden on individual branches, hence offers independent extensibility. The above example can be easily realized by:
\begin{lstlisting}
interface DrawableSafeDeck extends Drawable, SafeDeck {
  void draw() override Drawable {
    JFrame frame = new JFrame("Canvas");
    frame.setSize(600, 600);
    frame.getContentPane().setBackground(Color.red);
    frame.getContentPane().add(new Square(10,10,100,100)); ...
  }
}
// main program
((Drawable)new DrawableSafeDeck()).draw(); //calls the draw in DrawableSafeDeck
\end{lstlisting}

@@

In  Figure~\ref{fig:examplesmbody}, (a) is the base case for unintentional conflicts, namely the \wordfork{} inheritance. (b) uses overriding to explicitly merge the conflicting methods. (c) represents hierarchical overriding. Our model supports diamond inheritance and can deal with diamond problems. For example, (d) and (e) are two base cases of diamond inheritance in \MIM{} and the definition of each $C$ is rejected because $T$ is an ambiguous parent to $C$. One solution for diamond inheritance is to merge methods coming from different parents. (f) gives a common solution to the diamond as in Java or traits, which is to explicitly override $A.m$ and $B.m$ in $C$. And our calculus supports this kind of merging methods. In the last three examples, conflicting methods $A.m$ and $B.m$ should be viewed as
intentional conflicts, as they come from the same source $T$.
% Similar to many OO languages, the model also allows \textit{super
% method invocation} in a method body. The invocation
% \lstinline|super.T::m()| will ignore all the subtypes of
% \lstinline|T|, and only look at \lstinline|T| together with its
% super interfaces. It should behave the same as \lstinline|new
% T().m()| in principle.

>>>>

\begin{lstlisting}
interface DrawableSafeDeck extends Drawable, SafeDeck {}
new DrawableSafeDeck().shuffleAndDraw()
\end{lstlisting}

@@

% \begin{equation}\label{}
% \begin{split}
% A & = \frac{\pi r^2}{2} \\
%  & = \frac{1}{2} \pi r^2
% \end{split}
% \end{equation}
\vspace{1ex}
\begin{lstlisting}[style=reduction, escapechar=\%]
   %new DrawableSafeDeck()%.shuffleAndDraw()
-> %new DrawableSafeDeck().shuffle()%; new DrawableSafeDeck().draw()
-> ...
-> %new DrawableSafeDeck().draw()%
-> <<error: ambiguous call!!!>>
\end{lstlisting}

@@

\subsubsection{\MIM's solution} 
The essential problem is how to ensure that the correct method is
invoked. To solve this problem, \MIM{} uses a variant of method
dispatching that we call \textit{hierarchical dispatching}. In
hierarchical dispatching, both the static and dynamic type information 
are used to select the right method implementation.
%Inspired by C++, we have formalized this feature into a minimal core calculus,
%studied and proved the type soundness. It is concise and can readily be imported to other systems. 
%And in the formalization, we propose \textit{hierarchical dispatch} in our model
%as the default method lookup algorithm. 
During runtime, a method call
makes use of both the static type and the dynamic type of the
receiver, so it is a
combination of static and dynamic dispatching. Intuitively, the static type specifies one branch
to avoid ambiguity, and the dynamic type finds the most specific implementation on that branch.
To be specific, the following code is accepted by \MIM{}:
\begin{lstlisting}
interface Deck {
  void draw() {...}
  void shuffle() {...}
  void shuffleAndDraw() { this.shuffle(); this.draw(); }
}
interface Drawable {...}
interface SafeDeck extends Deck {...}
interface DrawableSafeDeck extends Drawable, SafeDeck {}
new DrawableSafeDeck().shuffleAndDraw() // SafeDeck.draw is called
\end{lstlisting}

\noindent The computation performed in \MIM{} is as follows:
\vspace{1ex}
\begin{lstlisting}[style=reduction, escapechar=\%]
   %new DrawableSafeDeck()%.shuffleAndDraw()
-> %((DrawableSafeDeck) new DrawableSafeDeck()).shuffleAndDraw()%
-> %((Deck) new DrawableSafeDeck()).shuffle()%; ((Deck) new DrawableSafeDeck()).draw()
-> ...
-> %((Deck) new DrawableSafeDeck()).draw()%
-> ... %// SafeDeck.draw%
\end{lstlisting}

@@

\subsubsection{\MIM{}'s solution} An additional feature of our model is \textit{hierarchical overriding}. It allows conflicting methods
to be overridden on individual branches, hence offers independent extensibility. The above example can be easily realized by:
\begin{lstlisting}
interface DrawableSafeDeck extends Drawable, SafeDeck {
  void draw() override Drawable {
    JFrame frame = new JFrame("Canvas");
    frame.setSize(600, 600);
    frame.getContentPane().setBackground(Color.red);
    frame.getContentPane().add(new Square(10,10,100,100)); ...
  }
}
((Drawable)new DrawableSafeDeck()).draw(); //calls the draw in DrawableSafeDeck
\end{lstlisting}

@@

In  Figure~\ref{fig:examplesmbody}, (a) is the base case for
unintentional conflicts, namely the \wordfork{} inheritance. (b) uses
overriding to explicitly merge the conflicting methods. (c) represents
hierarchical overriding.

Furthermore, our model supports diamond inheritance and can deal with diamond problems. For example, (d) and (e) are two base cases of diamond inheritance in \MIM{} and the definition of each $C$ is rejected because $T$ is an ambiguous parent to $C$. One solution for diamond inheritance is to merge methods coming from different parents. (f) gives a common solution to the diamond as in Java or traits, which is to explicitly override $A.m$ and $B.m$ in $C$. And our calculus supports this kind of merging methods. In the last three examples, conflicting methods $A.m$ and $B.m$ should be viewed as
intentional conflicts, as they come from the same source $T$.
% Similar to many OO languages, the model also allows \textit{super
% method invocation} in a method body. The invocation
% \lstinline|super.T::m()| will ignore all the subtypes of
% \lstinline|T|, and only look at \lstinline|T| together with its
% super interfaces. It should behave the same as \lstinline|new
% T().m()| in principle.
